Improved state-space models for inference about spatial and temporal variation in abundance from count data

### 2012-09-10 Dail_Mad_ext.tex
break

\abstract{
Models of population dynamics play a central role in theoretical and
applied ecology where they are used for purposes such as testing
hypotheses about density dependence and predicting species' responses
to future environmental change or conservation actions. Accounting for
process variation and observation error in such efforts is necessary
to achieve unbiasedness, and doing so is possible using recently
developed state-space models.
%for estimators of population change to be unbiased, and state-space
%models have been developed for this
%models are widely used for these purposes because they allow for
%estimating parameters of classical growth models, such as exponential,
%logistic, and Gompertz, while accounting for demographic stochasticity
%and observation error.
Conventional state-space models, however, have
three important limitations: (1) the parameters are not identifiable
in many common situations, (2) they do not admit spatial variation in
population dynamics, and (3) there is no clear interpretation of the
observation error. We demonstrate how each of these problems can be
resolved using a class of hierarchical models for spatially-replicated
time-series data recently proposed by \citet[Biometrics]{dail_madsen:2011}.
We expand this class of models to accommodate classical growth
models, zero-inflation, and random effects such as observer-specific
detection probabilities. We also present methods for forecasting
populaiton size under future environmental conditions. Implementation
of these ideas is possible using either frequentist or Bayesian
methods, and code to fit these models using the  \textbf{R} package
\texttt{unmarked} or
\textbf{JAGS} is also provided.
%is provided for doing so.
%Furthermore, we describe classical and
%Bayesian methods of parameter estimation, and we present methods of
%forecasting population size under future environmental
%conditions.
%These new developments will
%allow researchers to apply these methods to address important
%questions regarding the factors affecting spatial and temporal
%variation in abundance.
A simulation study was used to assess bias and accurrancy of parameter
estimators, and an analysis of %We demonstrate these developments by analyzing
%data from the
North American Breeding Bird Survey is presented for illustration.
}

@@

A simple state-space models can be described as follows.
Let $N_t$ be the abundance of a species during year $t$, for
$t=1,\hdots,T$, and let $X_t$ be
the observed data, which differs from $N_t$ due to observation error,
a random effect denoted $\zeta_t$. Temporal variation in $N_t$ is
modeled using a deterministic population growth model, $\mu(N_{t-1})$,
coupled with a stochastic model allowing
for random process variation. %, which we denote $\grave{o}_t$.
The deterministic model may be density-dependent, for example
logistic, or it might be density-independent, as in the case of
exponential growth, in which $\mu(N_{t-1}) = N_{t-1}e^{r}$ where $r$
is the intrinsic rate of increase. The full model can now be written
as:
\begin{subequations}
  \label{eq:ss1}
  \begin{align}
    N_1 &= X_1 \label{eq:ss1a} \\
    N_t &= \mu(N_{t-1}) + \grave{o}_{t-1} \quad \text{for} \; t=2,\hdots,T \label{eq:ss1b} \\
    X_t &= N_t + \zeta_t \qquad \qquad \;\, \text{for} \; t=1,\hdots,T \label{eq:ss1c}
  \end{align}
\end{subequations}
where $\grave{o}_t$ is the random effect allowing for process
variation unaccounted for by the deterministic model. In classical
state-space models, the two sets of random effects
are assumed to be i.i.d Gaussian deviates:
$\grave{o}_t \sim \mathrm{N}(0, \sigma)$ and
$\zeta_t \sim \mathrm{N}(0, \tau)$. It is also standard practice to
act as though there is no process variation associated with $N_1$, as
indicated by Eq~\ref{eq:ss1a}.
% indicates that no process variation is
%present in year 1. This assumption is often made
%due to parameter identifiability problems, which we will describe
%shortly. First, we turn our attention to the models for the random
%effects.

@@

Use of the Gaussian distribution for modeling random process variation
and observation error is motivated by convenience rather than
biology. Specifically, the Gaussian assumptions
allow for parameter estimation using the Kalman filter
\citep{dennis_etal:2006}, which is much more computationally efficient
than estimation methods when random effects are not Gaussian distributed
\citep{devalpine_hastings:2002}. The problem with this is that it
allows for negative values of the state-variable, which is
inconsistent with the observed data.
%The fact that both abundance ($N_t$) and the observed data ($X_t$) are
%typically non-negative integers suggests that it is inappropriate to
%model process variation and observation error using Gaussian
%distributions.
Nonetheless, efforts have been made to adhere to the
Gaussian assumptions by tranforming abundance to
density by replacing $N_t$ with $Y_t = \log(N_t/A)$ where $A$ is the
area surveyed. However, this transformation is problematic since
abundance may be zero and thus $Y_t = \log(0) = -\infty$. Generally,
zeros are replaced with some small number, although the
effect of this is rarely discussed.

@@

A more serious problem than the ones associated with the Gaussian
assumptions is that the parameters of the simple state-space models
such as Eq~\ref{eq:ss1} are not identifiable in many
circumstances \citep{polansky_etal:2009}. Recall that the
Eq.~\ref{eq:ss1} did not specify distributions for $N_1$; % or $X_1$;
however, $N_1$ is a random variable and hence there is uncertainty
that should be accounted for.
%are needed to fully specify the model.
To be more specific, a fully-specified state-space model requires at least
three probability distributions, which we represent
using bracket notation:
\begin{subequations}
  \label{eq:ss2}
  \begin{align}
    [N_1&|\bm{\theta}] \\ \label{eq:ss2b}
    [N_t&|N_{t-1},\bm{\Theta}] \quad \text{for} \; t=2,\hdots,T \\ \label{eq:ss2c}
    [X_t&|N_t,\bm{p}]  \qquad \; \text{for} \; t=1,\hdots,T
  \end{align}
\end{subequations}
where $\bm{\theta}$ are the process variation parameters for the
initial state, $\bm{\Theta}$ are the process variation parameters
influencing how abundance changes over time, and $\bm{p}$ are the
observation error parameters. The three lines of general model
described by Eq.~\ref{eq:ss2} correspond to the three equations shown
for the specific example in Eq~\ref{eq:ss1}.  However, in this case,
we allow $N_1$ to be a random variable. But what should its
distribution, $[N_1]$ be? And how could $\bm{\theta}$ be
estimated since there is only a single observation available?
One approach to estimating the $\bm{\theta}$ is to assume that the
population is in equlibrium, i.e. that the expected value of $N_t$ is
constant through time. Althought this makes the parameters
identifiable, assuming equlibrium defeats the objective of many
studies of population dynamics, namely determining why a population
varies over time.

@@

A fourth problem with these models is that they do not admit
spatial variation, which
%. Recent work has sought to remedy this deficiencyThis is not a fair criticism because this
%lies beyond the scope of traditional state-space models, but
%it does
restrict the utility of their since it is typically
impossible assess the impacts of factors such as habitat
fragmentation or climate change on population dynamics without
considering spatial variation in abundance. Indeed many populations are
regulated by spatial processes such as source-sink dynamics.

Several extensions of state-space models have been proposed to
overcome the limitations described
above. \citet{devalpine_hastings:2002} and \citet{kery_etal:2009}
described methods for fitting models with non-Gaussian distributions
for the process and observation errors. Observation models with more
intuitive interpretations, such as those that explicitly model
detection probability, have been proposed by
\citet{kery_etal:2009}. \citet{lele_etal:1998} and
\citet{kery_etal:2009} developed models allowing for inference about
spatial and temporal variation in abundance, and their developments
also resolved the problems of non-identifiability for the parameters
of the initial state at time $t=1$. Of these extensions, the work by
\citet{kery_etal:2009} is unique in that it addressed multiple limitations
simultaneously. However, their model did not include serial
depedence, which is a hallmark of population models. This limits the
utility of their model for making inferences about explicit population
processes. In contrast, recent work has sought to use state-space
models fot inference about population processes such as mortality and
recruitment. To do so, methods have been developed to combine multiple
sources of information, such as counts and mark-recapture data
\citep{besbeas_etal:2002, buckland_etal:2004,
  schaub_etal:2007}. Although these integrated
population models may be viewed as the gold standard in state-space
modeling, ecologists are not always so fortunate to have direct
information about vital rates, especially at large spatial scales.
Rather, count data are much more common, and
are produced by many of the largest monitoring programs in the world,
such as the North American Breeding Bird Survey
\citep[BBS;][]{robbins_etal:1986}.

@@

Maximum likelihood estimation of parameters in models with random
effects requires removing the random effects

@@

>>>>

\abstract{
Models of population dynamics play a central role in theoretical and
applied ecology where they are used for purposes such as testing
hypotheses about density dependence and predicting species' responses
to future environmental change or conservation actions. Accounting for
process variation and observation error in such efforts is necessary
to achieve unbiasedness, and doing so is possible using recently
developed state-space models.
%for estimators of population change to be unbiased, and state-space
%models have been developed for this
%models are widely used for these purposes because they allow for
%estimating parameters of classical growth models, such as exponential,
%logistic, and Gompertz, while accounting for demographic stochasticity
%and observation error.
Conventional state-space models, however, have
three important limitations: (1) the parameters are not identifiable
in many common situations, (2) they do not admit spatial variation in
population dynamics, and (3) there is no clear interpretation of the
observation error. We demonstrate how each of these problems can be
resolved using a class of hierarchical models for spatially-replicated
time-series data recently proposed by \citet[Biometrics]{dail_madsen:2011}.
We expand this class of models to accommodate classical growth
models, zero-inflation, and random effects such as observer-specific
detection probabilities. We also present methods for forecasting
population size under future environmental conditions. Implementation
of these ideas is possible using either frequentist or Bayesian
methods, and code to fit these models using the  \textbf{R} package
\texttt{unmarked} or
\textbf{JAGS} is also provided.
%is provided for doing so.
%Furthermore, we describe classical and
%Bayesian methods of parameter estimation, and we present methods of
%forecasting population size under future environmental
%conditions.
%These new developments will
%allow researchers to apply these methods to address important
%questions regarding the factors affecting spatial and temporal
%variation in abundance.
A simulation study indicated that bias was negligible and coverage nominal
and an analysis of North American Breeding Bird Survey is presented
for illustration.
}

@@

A simple state-space models can be described as follows.
Let $N_t$ be the abundance of a species during year $t$, for
$t=1,\hdots,T$, and let $X_t$ be
the observed data, which differs from $N_t$ due to observation error,
a random effect denoted $\zeta_t$. Temporal variation in $N_t$ is
modeled using a population growth model, $\mu(N_{t-1})$,
coupled with a stochastic model allowing
for random process variation. %, which we denote $\grave{o}_t$.
The deterministic model may be density-dependent, for example
logistic, or it might be density-independent, as in the case of
exponential growth, in which $\mu(N_{t-1}) = N_{t-1}e^{r}$ where $r$
is the intrinsic rate of increase. The full model can now be written
as:
\begin{subequations}
  \label{eq:ss1}
  \begin{align}
    N_1 &= X_1 \label{eq:ss1a} \\
    N_t &= \mu(N_{t-1}) + \grave{o}_{t-1} \quad \text{for} \; t=2,\hdots,T \label{eq:ss1b} \\
    X_t &= N_t + \zeta_t \qquad \qquad \;\, \text{for} \; t=1,\hdots,T \label{eq:ss1c}
  \end{align}
\end{subequations}
where $\grave{o}_t$ is the random effect allowing for process
variation unaccounted for by the deterministic model. In classical
state-space models, the two sets of random effects
are assumed to be i.i.d Gaussian deviates:
$\grave{o}_t \sim \mathrm{N}(0, \sigma)$ and
$\zeta_t \sim \mathrm{N}(0, \tau)$. It is also standard practice to
ignore process variation associated with $N_1$, as
indicated by Eq~\ref{eq:ss1a}.

@@

Use of the Gaussian distribution for modeling random process variation
and observation error is motivated by convenience rather than
biology. Specifically, the Gaussian assumptions
allow for parameter estimation using the Kalman filter
\citep{dennis_etal:2006}, which is much more computationally efficient
than estimation methods when random effects are not Gaussian distributed
\citep{devalpine_hastings:2002}. The problem with this is that it
allows for negative values of the state-variable, which is
inconsistent with the observed data.
%The fact that both abundance ($N_t$) and the observed data ($X_t$) are
%typically non-negative integers suggests that it is inappropriate to
%model process variation and observation error using Gaussian
%distributions.
Nonetheless, efforts have been made to adhere to the
Gaussian assumptions by tranforming abundance to
density by replacing $N_t$ with $Y_t = \log(N_t/A)$ where $A$ is the
area surveyed. However, this transformation is problematic since
local abundance may be zero and thus $Y_t = \log(0) = -\infty$. Generally,
zeros are replaced with some small number, although the
effect of this is rarely discussed.

@@

A more serious problem than the ones associated with the Gaussian
assumptions is that the parameters of the simple state-space models
such as Eq~\ref{eq:ss1} are not identifiable in many
circumstances \citep{polansky_etal:2009}. Recall that the
Eq.~\ref{eq:ss1} did not specify distributions for $N_1$; % or $X_1$;
however, $N_1$ is a random variable and hence there is uncertainty
that should be accounted for.
%are needed to fully specify the model.
To be more specific, a fully-specified state-space model requires at least
three probability distributions, which we represent
using bracket notation:
\begin{subequations}
  \label{eq:ss2}
  \begin{align}
    [N_1&|\bm{\theta}] \\ \label{eq:ss2b}
    [N_t&|N_{t-1},\bm{\Theta}] \quad \text{for} \; t=2,\hdots,T \\ \label{eq:ss2c}
    [X_t&|N_t,\bm{p}]  \qquad \; \text{for} \; t=1,\hdots,T
  \end{align}
\end{subequations}
where $\bm{\theta}$ are the process variation parameters for the
initial state, $\bm{\Theta}$ are the process variation parameters
influencing how abundance changes over time, and $\bm{p}$ are the
observation error parameters. The three lines of general model
described by Eq.~\ref{eq:ss2} correspond to the three equations shown
for the specific example in Eq~\ref{eq:ss1}.  However, in this case,
we allow $N_1$ to be a random variable. But what should its
distribution, $[N_1|\bm{\theta}]$ be? And how could $\bm{\theta}$ be
estimated since there is only a single observation available?
One approach to estimating the $\bm{\theta}$ is to assume that the
population is in equlibrium, i.e. that the expected value of $N_t$ is
constant through time. Althought this makes the parameters
identifiable, assuming equlibrium defeats the objective of many
studies of population dynamics, namely determining why a population
varies over time.

@@

A fourth problem with these models is that they do not admit
spatial variation. This reduces the scope of the inferences that can
be drawn from the models and it ignores the importance of space in
population regulation.
%, which
%restrict the utility of their since it is typically
%impossible assess the impacts of factors such as habitat
%fragmentation or climate change on population dynamics without
%considering spatial variation in abundance. Indeed many populations are
%regulated by spatial processes such as source-sink dynamics.

Several extensions of state-space models have been proposed to
overcome the limitations described
above. \citet{devalpine_hastings:2002} and \citet{kery_etal:2009}
described methods for fitting models with non-Gaussian distributions
for the process and observation errors. Observation models with more
intuitive interpretations, such as those that explicitly model
detection probability, have been proposed by
\citet{kery_etal:2009}. \citet{lele_etal:1998} and
\citet{kery_etal:2009} developed models allowing for inference about
spatial and temporal variation in abundance, and their developments
also resolved the problems of non-identifiability for the parameters
of the initial state at time $t=1$. Of these extensions, the work by
\citet{kery_etal:2009} is unique in that it addressed multiple limitations
simultaneously. However, their model did not include serial
depedence, which is a hallmark of population models. This limits the
utility of their model for making inferences about explicit population
processes. In contrast, recent work has sought to use state-space
models for inference about population processes such as mortality and
recruitment. To do so, methods have been developed to combine multiple
sources of information, such as counts and mark-recapture data
\citep{besbeas_etal:2002, buckland_etal:2004,
  schaub_etal:2007}. Although these integrated
population models may be viewed as the gold standard in state-space
modeling, ecologists are not always so fortunate to have direct
information about vital rates, especially at large spatial scales.
Rather, count data are much more common, and
are produced by many of the largest monitoring programs in the world,
such as the North American Breeding Bird Survey
\citep[BBS;][]{robbins_etal:1986}.

@@

Maximum likelihood estimation of parameters in models with random
effects requires removing the random effects by integration or
summation.

@@

\section{Forecasting Future Population Size}
